---
title: "The _checkpoint_ Package"
#author: "Stefan Schmidt"
#date: "7/24/2020"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Purpose

Main purpose of the _checkpoint_ package is __reproducibility in R__. This is
achieved by setting up a __frozen repository__. To do this _checkpoint_
modifies the R library path (pointing to a local directory) and changes the
_repos_ option to reference the specified MRAN snapshot.

The reproducibility issue is already solved on DaVinci -- MRAN snapshots
are currently used there.
Nevertheless DaVinci users can benefits from _checkpoint_ usage, because the
package provides a simple interface to modify project repository and library
path.


## Expected Benefits

### All contributed R packages are stored in one place

_checkpoint_ stores all contributed R packages in the _checkpoint_ library
path. By default this is a subdirectory of the users home directory
(~/.checkpoint/).

But this behaviour can be modified and _checkpoint_ advised to choose a
different local directory, which is especially usefull when doing GxP
related work where all used contributed packages have to be stored in
a special repository like ClearCase.

Let us assume the following case:

* a DaVinci users work is GxP related and
* all packages used by the project have to be stored in ClearCase

In this situation the _checkpoint_ package comes in handy. By using the
`checkpoint()` function it is relatively easy to not only define which MRAN
snapshot to use, but also to configure your library folder to be stored in
your project directory. This is achieved by setting the `checkpointLocation`
parameter:

```{r checkpointLocation, eval=FALSE}
library(checkpoint)
checkpoint('2019-11-22', checkpointLocation='~/project_dir/')
```

In this case it is also an advantage that the R library path
is shrunk down to the location of your R installation and the _checkpoint_
library folder only.
All used packages (except base R and recommended packages) are then stored in
your project directory and can be checked-in to ClearCase.


### R version pre-defined in your script

As a precaution even your R version can be determined when calling the
`checkpoint()` function:

```{r R.version, eval=FALSE}
library(checkpoint)
checkpoint('2019-11-22', R.version='3.6.1', checkpointLocation='~/project_dir/')
```

This will result in an error message when you are not in your proper
environment (for whatever reasons) or the environment has changed to a
different R version.


## Possible Issues

### Packages are re-installed to your local _checkpoint_ library

DaVinci provides many contributed packages. Using _checkpoint_ even packages
that are already installed on DaVinci will be re-installed to your _checkpoint_
directory.

This has pros and cons.

#### PROs:

* only a __limited number of packages__ namely the ones that are used in your
project will be contained in your _checkpoint_ library.
Especially GxP related work where all used source code (including used
packages) has to be checked in to ClearCase is then easy to reproduce. The only
component that is then not checked in to ClearCase is the R version used.
And even this information can be easily recorded (using the _R.version_
parameter).

#### CONs:

* local re-installation of packages probably has to be covered by some kind of
__installation testing__ (even when the packages are available pre-installed
on DaVinci)


### Default location of _checkpoint_ library path

The default location for the _checkpoint_ library path is the users home
directory and there is no way for _checkpoint_ find a library path that
differs from the default.



## Background Information

### CRAN, MRAN and checkpoint

#### CRAN

CRAN follows a rolling release concept. R packages are always provided in the
latest version.

This leads to issues:

* custom-built R scripts can break in the future when used packages are updated
* additional package installations can break dependencies of another already
installed package
* results are not reproducible since the environment in which they were
obtained changes over time


#### MRAN

The Microsoft R Application Network (MRAN) solves part of these issues by the
creation of [daily CRAN snapshots](https://cran.microsoft.com/snapshot/)
that are provided back to the R community as a service.
Every day since September 2014 at midnight UTC a snapshot of the CRAN main
server in Austria is rsynced to MRAN.


#### checkpoint

The `checkpoint` package makes it possible to install package versions from one
of the MRAN specific date in the past referencing the MRAN snapshots.




## FAQ

### What snapshot dates are recommended?

A suitable snapshot should meet the following criteria:

* _availabiltiy_ - there is not for every day a snapshot available on MRAN.
Valid snapshot dates can be retrieved via `getValidSnapshots()`.
* _intended R version_ - to prevent incompatibilities a snapshot for the used
R version should be chosen.
* _latest bugfixes included_ - in case the R version is fixed (like on
DaVinci), it is reasonable to choose a later snapshot to include a maximum of
package bugfixes available for your R version. 





## What happens in the background

### Library Path

The _checkpoint_ package modyfies the R library path.
Using _checkpoint('2019-11-22')_ the output of `.libPaths()` changes from:

```
R 3.6.1> .libPaths()
[1] "/CHBS/apps/EB/software/R-bundle-Novartis/0.2-gomkl-2019a-R-3.6.1"    
[2] "/CHBS/apps/EB/software/R-bundle-Bioconductor/3.9-gomkl-2019a-R-3.6.1"
[3] "/CHBS/apps/EB/software/ncdf4/1.16.1-gomkl-2019a-R-3.6.1"             
[4] "/CHBS/apps/EB/software/R/3.6.1-gomkl-2019a/lib64/R/library"
```
to:
```
R 3.6.1> .libPaths()
[1] "/home/schmis1m/.checkpoint/2019-11-22/lib/x86_64-pc-linux-gnu/3.6.1"
[2] "/home/schmis1m/.checkpoint/R-3.6.1"                                 
[3] "/CHBS/apps/EB/software/R/3.6.1-gomkl-2019a/lib64/R/library"
```

As a side effect the number of available packages has also changed from 1'530
to only 705 (plus user installed packages). Since the _checkpoint_ package
itself is part of module 'R-bundle-Novartis' it is now no longer available:

```{r inst_pkg}
ip <- as.data.frame(installed.packages())
any(ip$Package == 'checkpoint')
```


### Package Repositories

Running the _checkpoint_ function on DaVinci makes package repositories change
from :
```
R 3.6.1> options('repos')
$repos
                                                   CRAN 
      "https://cran.microsoft.com/snapshot/2019-11-22/" 
                                               BioCsoft 
           "https://bioconductor.org/packages/3.9/bioc" 
                                                BioCann 
"https://bioconductor.org/packages/3.9/data/annotation" 
                                                BioCexp 
"https://bioconductor.org/packages/3.9/data/experiment" 
```
to:
```
R 3.6.1> options('repos')
$repos
[1] "https://mran.microsoft.com/snapshot/2019-11-22"
```

## Rollback with `unCheckpoint()`

Executing `unCheckpoint()` will restore the library path back to:


```
R 3.6.1> .libPaths()
[1] "/CHBS/apps/EB/software/R-bundle-Novartis/0.2-gomkl-2019a-R-3.6.1"    
[2] "/CHBS/apps/EB/software/R-bundle-Bioconductor/3.9-gomkl-2019a-R-3.6.1"
[3] "/CHBS/apps/EB/software/ncdf4/1.16.1-gomkl-2019a-R-3.6.1"             
[4] "/CHBS/apps/EB/software/R/3.6.1-gomkl-2019a/lib64/R/library"
```

But package repositories will continue to link to the MRAN snapshot only.




## Open Questions

### How can non-CRAN packages be installed reproducibly?

_checkpoint_ internally uses `utils::install.packages()` for package
installation. Can packages from other repositories (BioConductor, GitHub) also
be installed in a reproducible way?

### Are a packages system requirements a problem?

Probably yes, if a new manually installed packages expects some system
libraries to be available that have not already been installed by one of the
DaVinci pre-installed packages.


## Notes

### checkpoint 1.0 ante portas

In April 2020 Hong Ooi, the package maintainer, announced
[checkpoint version 1.0 beta](https://blog.revolutionanalytics.com/2020/04/checkpoint-10-now-in-beta.html)
to be available on GitHub.



## Download Statistics

CRAN task view packages for
[reproducible research](https://cloud.r-project.org/web/views/ReproducibleResearch.html)
(section: package reproducibility) are:

* checkpoint
* miniCRAN
* packrat
* rbundler
* renv
* switchr

Download statistics for these packages look like this:

```{r dl_stats, echo=FALSE}
library(ggplot2)
library(dlstats)
#library(checkpoint)

#checkpoint('2019-11-22',
#           R.version = '3.6.1',
#           checkpointLocation = '/home/schmis1m/wrk/studio/r_pkg_evaluations/checkpoint/')


dls <- cran_stats(c("checkpoint", "miniCRAN", "packrat", "rbundler", "renv", "switchr"))

if (!is.null(dls)) {
  head(dls)
  ggplot(dls, aes(end, log(downloads), group = package, color = package)) +
    geom_line() + geom_point(aes(shape = package))
}

```

